---
title: "Exploring the Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Brian Waismeyer"
date: "Sunday, May 24, 2015"
output: pdf_document
---
## Overview
In this report we will explore the ToothGrow dataset included in the default
R `datasets` package. This dataset is based on a research study with 10 guinea
pigs who were given a vitamin C supplement.

The researchers varied how the supplement was delivered and in what dose. They 
measured guinea pig tooth length for all 10 guinea pigs for each dose/delivery
combination (2 delivery types * 3 doses * 10 guinea pigs = 60 records total).

More information about this dataset is available 
[here](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html).

We will investigate and test the hypotheses that:

1. Tooth length is associated with the manner in which the vitamin C supplement
was delivered.
2. Tooth length is associated with the size of vitamin C dose given to a guinea
pig.


## Load Supporting Resources
We take a moment to load any packages we want to use in this report. These need
to be installed and loaded in your R environment if you want to reproduce
the report in full.

```{r}
library(dplyr)
library(ggplot2)
library(ggthemes)
```

## Load and Explore the Data
We load the data and visualize the key variables of interest.

```{r}
# load the data and then assign the dataset to a shorter name for convenience
# (and in case we want to return to the original data for any reason)
data(ToothGrowth)
tg <- ToothGrowth

# inspect the dataframe
str(tg)

# add a factored version of dose for cases where we might want to treat this
# as a grouping variable (e.g. plotting)
tg$dose_level <- factor(tg$dose)

# observe the mean tooth length and standard deviation for each delivery type
sum_supp <- summarise(group_by(tg, supp), 
                      mean_length = mean(len),
                      sd_length = sd(len))

# observe the mean tooth length and standard deviation for each dose combination
sum_dose <- summarise(group_by(tg, dose_level), 
                      mean_length = mean(len),
                      sd_length = sd(len))

# for each dose/delivery combination
sum_combo <- summarise(group_by(tg, supp, dose_level), 
                       mean_length = mean(len),
                       sd_length = sd(len))

# print the summaries along with a visualization of each
sum_supp
ggplot(sum_supp, aes(x = supp, y = mean_length)) +
    geom_bar(stat = "identity") +
    labs(title = "Mean Tooth Length by Delivery Type") +
    theme_fivethirtyeight()

sum_dose
ggplot(sum_dose, aes(x = dose_level, y = mean_length)) +
    geom_bar(stat = "identity") +
    labs(title = "Mean Tooth Length by Dose") +
    theme_fivethirtyeight()

sum_combo
ggplot(sum_combo, aes(x = supp, y = mean_length, fill = dose_level)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = "Mean Tooth Length by Delivery Type and Dose") +
    theme_fivethirtyeight()

```


## Hypothesis Testing
Here we perform the calculations needed to assess whether our observed data
support our hypotheses or are insufficient to draw clear conclusions.

The most appropriate approach would be test the overall delivery type ~ dosage
relationship using a multi-way ANOVA. However, assignment directions dictate
that we only make use of techniques discussed in the current course.

As a result, we will simply test the hypotheses associated with our main effects
and we will simply employ t-tests to explore the relevant pairwise comparisons.

### Hypothesis 1: Delivery Type ~ Tooth Length
Is tooth length associated with the manner in which the vitamin C supplement
was delivered?

```{r}
# we use a conservative t-test: two-sided, assuming unequal variance
# since we don't have pairing information, we treat the groups as independent
t.test(len ~ supp,
       data = tg,
       alternative = "two.sided",
       var.equal = FALSE
       )
```

The null hypothesis we are tested is that the difference between our sample
means is 0 (aka - the samples have the same means).

The observed p-value suggests marginal significance - we don't make the classic
p < 0.05 cut-off, but we're close. We can see this marginal significance echoed
in the 95% confidence interval - the majority of likely distributions of
differences between the means include values larger than 0.

I would treat this as weak evidence that the delivery type may be associated
with tooth length, with OJ being associated with longer teeth.


### Hypothesis 2: Dosage ~ Tooth Length
Is tooth length associated with the size of the vitamin C dosage?

Ideally, we would test this with an ANOVA. However, per assignment directions,
we will restrict our analysis to those presened in the course and will instead
treat the results as being three pairwise comparisons.

```{r}
# as before, we use conservative t-tesst: two-sided, assuming unequal variance
# since we don't have pairing information, we treat the groups as independent

# 0.5 v 1.0
tg_05_10 <- tg[tg$dose_level == "0.5" | tg$dose_level == "1", ]
t.test(len ~ dose_level,
       data = tg_05_10,
       alternative = "two.sided",
       var.equal = FALSE
       )

# 0.5 v 2.0
tg_05_20 <- tg[tg$dose_level == "0.5" | tg$dose_level == "2", ]
t.test(len ~ dose_level,
       data = tg_05_20,
       alternative = "two.sided",
       var.equal = FALSE
       )

# 1.0 v 2.0
tg_10_20 <- tg[tg$dose_level == "1" | tg$dose_level == "2", ]
t.test(len ~ dose_level,
       data = tg_10_20,
       alternative = "two.sided",
       var.equal = FALSE
       )

```

All comparisons were statistically significant. That is, the observed p-values
and confidence intervals suggest that it is unlikely the true distribution of
differences between sample means includes "0" (no differences between the 
samples) for any of the comparisons.

This supports the hypothesis that there is an association between dosage and
tooth length. The magnitude of the dosages and the observed tooth lengths
suggest that tooth length tends to increase with dosage.


## Assumptions and Conclusions
For purpose of this report, we have assumed that it is acceptable to treat our
comparison groups as independent. Given that we know the data came from the 
same guinea pigs (i.e., the comparison are within-group rather than between),
this is not ideal. However, independent groups t-tests tend to have less
power than paired t-tests and we were conservative about the kinds of t-tests
we ran.

To avoid an explosion in the number of pairwise comparison we needed to conduct,
we also restricted our analyses to only investigating main effects. It is
possible, however, that interactions between independent variables can make
interpreting main effects problematic. Thus, we also assumed that any main
effects we observed would hold even in the presence of possible delivery type ~
dosage interaction.

With the above assumptions in mind, we observed weak evidence that delivery type
is associated with tooth length (OJ being associated with longer teeth than 
ascorbic acid) and reasonable evidence that dosage size is
associated with tooth length (larger doses being associated with longer teeth.

It would appear that delivery type and dosage should be considered if trying
to understand or influence tooth length in guinea pigs.